---
title: "Template for Joy Data Analysis"
output: rmdformats::readthedown
css: custom.css
vignette: >
  %\VignetteIndexEntry{joy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE, 
  eval = TRUE,
  cache = FALSE,
  dev.args = list(bg = "transparent"),
  out.width = 600
)

knitr::knit_hooks$set(output = multimedia::ansi_aware_handler)
options(crayon.enabled = TRUE)

library(tidyverse)
options(
  ggplot2.discrete.colour = c("#9491D9", "#F24405", "#3F8C61",  "#8C2E62", "#F2B705", "#11A0D9"),
  ggplot2.discrete.fill = c("#9491D9", "#F24405", "#3F8C61", "#8C2E62", "#F2B705", "#11A0D9"),
  ggplot2.continuous.colour = function(...) scale_color_distiller(palette = "Spectral", ...),
  ggplot2.continuous.fill = function(...) scale_fill_distiller(palette = "Spectral", ...)
)

th <- theme_classic() +
  theme(
    panel.background = element_rect(fill="transparent"),
    strip.background = element_rect(fill="transparent"),
    plot.background = element_rect(fill="transparent", color=NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.background = element_rect(fill="transparent"),
    legend.box.background = element_rect(fill="transparent"),
    legend.position = "bottom"
  )

theme_set(th)
```

```{r libraries}
library(multimedia)
library(tidyverse)
library(ggraph)
```

This vignette gives a quick introduction to the `multimedia` package using a
simulated dataset that is supposed to look a little like the BeWell data. The
basic question is -- we know that meditation can reduce depression and anxiety
symptoms, is it possible that microbiome changes might be playing a role? In the
language of mediation analysis, does the microbiome mediate changes in PHQ
scores?

```{r example_data}
demo_joy()
```

For mediation analysis, we'll want to distinguish between the different variable
types. This data structures takes care of that. Notice that we can specify
groups using tidyselect-style notation.

```{r setup}
exper <- mediation_data(demo_joy(), "PHQ", "treatment", starts_with("ASV"))
exper
```

This is the main estimation function. By default, we fit a separate linear
regression model for each mediation and outcome variable.

```{r}
model <- multimedia(exper) |>
  estimate(exper)

model
```

The `edges` slot tracks all the variable relationships.

```{r}
ggraph(model@edges) +
  geom_edge_link(arrow = arrow()) +
  geom_node_label(aes(label = name, fill = node_type))
```

We can look at the estimated mediation and outcome models. (Perhaps we should
write simpler accessors?)

```{r}
model@mediation@estimates
model@outcome@estimates
```

Now that we've coupled the mediation and outcome models, we can propogate
predictions and samples through them. That is, we can define certain
configurations of the treatment (and pretreatments, if we have them) and then
use the fitted models to simulate new mediation and outcome samples. By default,
it will sample at the template data that was used to fit the model, just like
the `predict` method for `lm`.

```{r}
sample(model)
predict(model)
```

Things get more interesting when we sample at new treatment and pretreatment
configurations. We need to be careful with our accounting, because we want the
flexibility to provide different combinations of treatments to different sets of
edges. For example, we may want to imagine that the edge for one particular
mediator was set to treatment while all others were left at control. The example
below has one sample with this kind of configuration and three others that keep
all edges at control.

```{r}
t_mediator <- model@treatments[c(2, rep(1, 3)), ]
t_outcome <- model@treatments[rep(1, 4), ]
profile <- setup_profile(model, t_mediator, t_outcome)
sample(model, profile = profile)
predict(model, profile = profile)
```

We can also contrast the predictions and samples under different profiles.

```{r}
profile_control <- setup_profile(model, model@treatments[rep(1, 4), ], t_outcome)
contrast_predictions(model, profile, profile_control)
contrast_samples(model, profile, profile_control)
```

# Effect Estimates

It's a small step from contrasting different configurations to asking for the
direct and indirect treatments effects. The direct effect is defined as the
average of $\hat{Y}\left(\hat{M}\left(t'\right), 1\right) - \hat{Y}\left(\hat{M}\left(t'\right), 0\right)$ 
across mediator treatment effects $t'$. The hats mean that we use the predicted
values from the mediation and outcome values. I've distinguished between
"overall" and "pathwise" indirect effects because we're working with
high-dimensional mediators. In the overall effect, we toggle treatment/control
status for incoming edges to all mediators. In pathwise indirect effects, we
toggle only the treatment going into one mediator.

```{r}
direct_effect(model, exper)
indirect_overall(model, exper)
indirect_pathwise(model, exper)
```

So far, we've done everything using just linear models. We could actually have
computed all these effects just by looking at parameter estimates. What's nice
is that we can plug in many differnet kinds of mediation or outcome models. The
package already includes interfaces to the logistic-normal multinomial, sparse
regression with glmnet, random forests with ranger, and bayesian models with
brms. It's also not too difficult to extend to new model types (we should add a
vignette). Here's an example of everything we did above but for glmnet. The fact
that all the estimates are 0 is a good thing -- there are no real effects in the
simulated data.

```{r}
model <- multimedia(exper,  glmnet_model(lambda = .1)) |>
  estimate(exper)

direct_effect(model, exper)
indirect_overall(model, exper)
indirect_pathwise(model, exper)
```

# Inference

Effect estimates are rarely enough on their own. We need some uncertainty
assessments to set appropriate expectations. The most straightforward approach
is to use the bootstrap.  Each element in `funs` will get its own data.frame
with the bootstrap distribution for that estimator.

```{r}
funs <- list(direct_effect = direct_effect)
bootstrap(model, exper, funs, B = 10)
```

We can also generate synthetic nulls to calibrate selection sets. The third
argument says which set of edges we want to remove under the null. In this case
we will generate synthetic null data where there is known to be no relationship
between the mediators and outcome. The fourth argument says which effect
estimates we should evaluate. We then fit the full model on both the original
and the synthetic null data. We can define false discovery rate thresholds by
ranking estimates across the two data sets. If we see many null effects mixed in
among the strong effects in real data, we know to trust only the very strongest
real effects (if any).

```{r}
contrast <- null_contrast(model, exper, "M->Y", indirect_pathwise)
fdr <- fdr_summary(contrast, "indirect_pathwise", 0.05)
fdr
```